\thetitle

### 2021-03-05 writing/challenges.tex
Fix citation style

While at first glance the method of section~\ref{sec:why-rational} seems
straightforward to implement and test, three challenges remain. The first is
comparing the effect of blame on the rational programmer across three different
mechanisms; second is the collection of a large number of representative
debugging scenarios; and third is finding a proper sampling strategy for the
huge space of possibilities. A coincidental challenge is the disparity of the
implementations of gradually typed languages. To eliminate this variable, the
authors use Racket, which is thus far the only language in which all three major
soundness mechanisms are available in a robust and comparable manner:
Typed Racket implements Natural, Shallow Racket~\cite{ttt21}
Transient, and plain Racket for Erasure.

@@

An alternative is to {\em generate a large and representative corpus of type
mistakes \/} using mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis}. Conventional mutation analysis is useless, however.  Mutation
analysis traditionally aims to inject bugs that challenge test suites, and it
discards those that yield ill-typed mutants as \emph{incompetent}. Indeed,
mutation analysis frameworks are fine-tuned to avoid them, and yet, it is
precisely those mutators that are needed for evaluating blame assignment strategies.

Based on a related experience, Gopinath and Walkingshaw~\cite{gw-mutation} write,
``existing mutation frameworks \ldots\ do not generate the kinds of mutations
needed to best evaluate type annotations'' and, worse, ``it is surprisingly
difficult to come up with mutants that actually describe subtle type faults.''
While the goal of their work---to evaluate the quality of types in
Python---is unrelated to the one here, the mechanism is related. And their
judgment confirms the experience of the authors.

>>>>

While at first glance the method of section~\ref{sec:why-rational} seems
straightforward to implement and test, three challenges remain. The first is
comparing the effect of blame on the rational programmer across three different
mechanisms; second is the collection of a large number of representative
debugging scenarios; and third is finding a proper sampling strategy for the
huge space of possibilities. A coincidental challenge is the disparity of the
implementations of gradually typed languages. To eliminate this variable, the
authors use Racket, which is thus far the only language in which all three major
soundness mechanisms are available in a robust and comparable manner:
Typed Racket implements Natural, Shallow Racket~\citep{ttt21}
Transient, and plain Racket for Erasure.

@@

An alternative is to {\em generate a large and representative corpus of type
mistakes \/} using mutation analysis~\citep{lipton1971fault, demillo1978hints,
jia2011analysis}. Conventional mutation analysis is useless, however.  Mutation
analysis traditionally aims to inject bugs that challenge test suites, and it
discards those that yield ill-typed mutants as \emph{incompetent}. Indeed,
mutation analysis frameworks are fine-tuned to avoid them, and yet, it is
precisely those mutators that are needed for evaluating blame assignment strategies.

Based on a related experience, Gopinath and Walkingshaw~\citep{gw-mutation} write,
``existing mutation frameworks \ldots\ do not generate the kinds of mutations
needed to best evaluate type annotations'' and, worse, ``it is surprisingly
difficult to come up with mutants that actually describe subtle type faults.''
While the goal of their work---to evaluate the quality of types in
Python---is unrelated to the one here, the mechanism is related. And their
judgment confirms the experience of the authors.

### 2021-03-05 writing/conclusion.tex
Fix citation style

The paper does {\em not\/} address a problem in the gradually typed world that
was pointed out early on by practical researchers~\cite{incorrect-ts,
sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
attention~\cite{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations
themselves.  Developers use gradual typing to move an untyped code base into the
typed realm, and to this end, they need typed APIs for the vast repositories of
already-existing libraries. Instead of converting the libraries themselves,
language implementors merely create typed facade modules that import untyped
functions and export them with type annotations, like {\tt typed-pack-lib} in
figure~\ref{fig:example}.  With those facades, the compiler can easily
type-check typed modules. The problem is that these retroactive additions of
types to a library may result from a misunderstanding of the code. \emph{Any
retroactively ascribed type may thus be a mistake itself.}

>>>>

The paper does {\em not\/} address a problem in the gradually typed world that
was pointed out early on by practical researchers~\citep{incorrect-ts,
sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
attention~\citep{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations
themselves.  Developers use gradual typing to move an untyped code base into the
typed realm, and to this end, they need typed APIs for the vast repositories of
already-existing libraries. Instead of converting the libraries themselves,
language implementors merely create typed facade modules that import untyped
functions and export them with type annotations, like {\tt typed-pack-lib} in
figure~\ref{fig:example}.  With those facades, the compiler can easily
type-check typed modules. The problem is that these retroactive additions of
types to a library may result from a misunderstanding of the code. \emph{Any
retroactively ascribed type may thus be a mistake itself.}

### 2021-03-05 writing/discussion.tex
Fix citation style

The results of the experiment suggests a number of high-level conclusions about
blame strategies in the gradual typed world.  First, run-time type checks have a
large positive impact, regardless of whether these checks assign blame or throw
plain exceptions.  Second, error messages with blame assignments are more
helpful than those without. The results also indicate, though, that blame is not
critical in a majority of cases, and therefore they suggest investigating
whether blame tracking is worth the performance cost.  Third, the Natural
approach fares better than the Transient approach, but only by a small margin.
Since Natural offers complete and sound path-based blame while Transient offers
incomplete but sound heap-based blame~\cite{gfd-oopsla-2019}, the results call
for a study concerning the relative strengths of the two models of blame.
Fourth, given that Transient's sound but shallow run-time type checks do not
seem to hamper debugging, a version of the Natural semantics of Typed Racket
that disables some of its wrappers may offer an answer to the well-known
performance issues of Natural~\cite{gtnffvf-jfp-2019}.  Fifth, given that both
modes of the Transient rational programmer are equally successful and that
Transient-last produces shorter trails, Reticulated Python could limit the size
of blame sequences to the last few entries to reduce its serious performance
problems (see below).

>>>>

The results of the experiment suggests a number of high-level conclusions about
blame strategies in the gradual typed world.  First, run-time type checks have a
large positive impact, regardless of whether these checks assign blame or throw
plain exceptions.  Second, error messages with blame assignments are more
helpful than those without. The results also indicate, though, that blame is not
critical in a majority of cases, and therefore they suggest investigating
whether blame tracking is worth the performance cost.  Third, the Natural
approach fares better than the Transient approach, but only by a small margin.
Since Natural offers complete and sound path-based blame while Transient offers
incomplete but sound heap-based blame~\citep{gfd-oopsla-2019}, the results call
for a study concerning the relative strengths of the two models of blame.
Fourth, given that Transient's sound but shallow run-time type checks do not
seem to hamper debugging, a version of the Natural semantics of Typed Racket
that disables some of its wrappers may offer an answer to the well-known
performance issues of Natural~\citep{gtnffvf-jfp-2019}.  Fifth, given that both
modes of the Transient rational programmer are equally successful and that
Transient-last produces shorter trails, Reticulated Python could limit the size
of blame sequences to the last few entries to reduce its serious performance
problems (see below).

### 2021-03-05 writing/ideas.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions
Fix citation style

The rational programmer exploits blame information to figure out wih. Therefore the way to understand its behavior 
is to analyze the migratory path it follows through the 
\emph{configuration space} of the program. The space
consists of variants of the program, dubbed \emph{configurations}, that 
differ in their type annotations. 
Herein we build on top of Typed Racket's migratory typing
approach,\footnote{The method can be adapted to so-called true gradual typing
systems~\cite{svcb-snapl-2015} in a straightforward manner.}, hence, each configuration differs from another
in which of its components are typed and which are not. Formally, we can
represent a program $\system$ as a set of of components
$\set{\component}$ and each configuration $\conf$ of $\system$ is a subset of 
$\set{\component}$
that contains exactly the components of  $\system$ that are typed in
$\conf$. Naturally the configurations of  $\system$ are ordered by the subset relation and
thus form a lattice $\lattice{\system}$ with size $2^{\size{\system}}$.
The bottom configuration of $\lattice{\system}$ is always $\emptyset$ and
the top element is $\system$ itself. In between there are all the
mixed-type configurations. If we assume that  component $\component$ of $\system$
contains the type error, any configuration $\conf$ that
includes $\component$ is ill-typed. Any other configuration $\conf$ may blame 
a $\component_b$, which we denote with $\blame{\conf}$. Let a blametrail
be a sequence of configurations $\conf_0,...\conf_n$ where 
for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \blame{\conf_i}$. That is a blametrail
corresponds to a succession of a configurations that start with a
configuration $\conf_0$, dubbed the \emph{root of the blametrail}, and 
each configuration is derived from its previous one by typing the
component that the latter blames. A rational programmer is one that
migrates the variant of a program $\system$ that corresponds to
configuration $\conf_0$ along the blametrail that starts at $\conf_0$.
We say a blametrail is final iff its last configuration, $\conf_n$
doesn't blame a component, $\blame{\conf_i} = \emptyset$. We call a
blametrail \emph{successful} iff it is final and its last configuration
$\conf_n$ contains the faulty component of $\system$, $\faulty{\system}
\in \conf_n$.

>>>>

The rational programmer exploits blame information to figure out wih. Therefore the way to understand its behavior 
is to analyze the migratory path it follows through the 
\emph{configuration space} of the program. The space
consists of variants of the program, dubbed \emph{configurations}, that 
differ in their type annotations. 
Herein we build on top of Typed Racket's migratory typing
approach,\footnote{The method can be adapted to so-called true gradual typing
systems~\citep{svcb-snapl-2015} in a straightforward manner.}, hence, each configuration differs from another
in which of its components are typed and which are not. Formally, we can
represent a program $\system$ as a set of of components
$\set{\component}$ and each configuration $\conf$ of $\system$ is a subset of 
$\set{\component}$
that contains exactly the components of  $\system$ that are typed in
$\conf$. Naturally the configurations of  $\system$ are ordered by the subset relation and
thus form a lattice $\lattice{\system}$ with size $2^{\size{\system}}$.
The bottom configuration of $\lattice{\system}$ is always $\emptyset$ and
the top element is $\system$ itself. In between there are all the
mixed-type configurations. If we assume that  component $\component$ of $\system$
contains the type error, any configuration $\conf$ that
includes $\component$ is ill-typed. Any other configuration $\conf$ may blame 
a $\component_b$, which we denote with $\blame{\conf}$. Let a blametrail
be a sequence of configurations $\conf_0,...\conf_n$ where 
for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \blame{\conf_i}$. That is a blametrail
corresponds to a succession of a configurations that start with a
configuration $\conf_0$, dubbed the \emph{root of the blametrail}, and 
each configuration is derived from its previous one by typing the
component that the latter blames. A rational programmer is one that
migrates the variant of a program $\system$ that corresponds to
configuration $\conf_0$ along the blametrail that starts at $\conf_0$.
We say a blametrail is final iff its last configuration, $\conf_n$
does not blame a component, $\blame{\conf_i} = \emptyset$. We call a
blametrail \emph{successful} iff it is final and its last configuration
$\conf_n$ contains the faulty component of $\system$, $\faulty{\system}
\in \conf_n$.

### 2021-03-05 writing/introduction.tex
Fix citation style

Theoreticians of gradual typing have focused on blame theorems from the
very beginning~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
[components]\footnote{The original authors got this word wrong.} can't be
blamed'' turned the theorem into a slogan~\cite{wf-esop-2009}. Academic
systems (Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017,
vksb-dls-2014} and Typed
Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010})
come with sophisticated checking and blame assignment strategies (see
sec.~\ref{sec:landscape}). Their academic creators embrace the idea that
blame can help practicing programmers find impedance mismatches, that is,
disagreements between the type ascriptions of a software system and its
behavior.

@@

The paper's {\it second contribution\/} is a set of {\em results from
applying the evaluation method to three distinct blame assignment
strategies in approximately 72,000 different scenarios\/} (see
sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
tracking of typed/untyped boundaries~\cite{vss-popl-2017}; (2)~{\it
Natural\/}, i.e. Typed Racket's use of a higher-order contract system and
its blame assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
the approach of industrial systems, which forgo blame in favor of error
messages from the safety checks of the underlying language. The results
(see sec.~\ref{sec:discussion}) are at least somewhat surprising.  In
principle they validate the conjectures behind the work of theoreticians.
First, a good blame assignment strategy helps with the search for
impedance mismatches between the specified types and the behavior of
untyped components.  Second, Natural's wrapper-based blame tracking is
more useful than Transient's ``collective blame'' tracking algorithm,
which in turn is superior to Erasure. {\em But\/}, the application of the
method also indicates problems with the expectations of theoreticians. The
first problem concerns the difference between the methods; neither is {\it
Natural\/} vastly more useful than {\it Transient\/} nor are the two
blame assignment methods highly superior to {\it Erasure\/}.  The second
one is that the cost of Transient's blame can be huge.  In turn, these
problems suggest that, on one hand, the existing theory does not predict
practice properly, and on the other hand, the existing practice may need
additional experiments to collect more data.

>>>>

Theoreticians of gradual typing have focused on blame theorems from the
very beginning~\citep{mf-toplas-2009, tf-dls-2006}. ``Well-typed
[components]\footnote{The original authors got this word wrong.} can't be
blamed'' turned the theorem into a slogan~\citep{wf-esop-2009}. Academic
systems (Reticulated Python~\citep{vsc-dls-2019, vss-popl-2017,
vksb-dls-2014} and Typed
Racket~\citep{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010})
come with sophisticated checking and blame assignment strategies (see
sec.~\ref{sec:landscape}). Their academic creators embrace the idea that
blame can help practicing programmers find impedance mismatches, that is,
disagreements between the type ascriptions of a software system and its
behavior.

@@

The paper's {\it second contribution\/} is a set of {\em results from
applying the evaluation method to three distinct blame assignment
strategies in approximately 72,000 different scenarios\/} (see
sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
tracking of typed/untyped boundaries~\citep{vss-popl-2017}; (2)~{\it
Natural\/}, i.e. Typed Racket's use of a higher-order contract system and
its blame assignment~\citep{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
the approach of industrial systems, which forgo blame in favor of error
messages from the safety checks of the underlying language. The results
(see sec.~\ref{sec:discussion}) are at least somewhat surprising.  In
principle they validate the conjectures behind the work of theoreticians.
First, a good blame assignment strategy helps with the search for
impedance mismatches between the specified types and the behavior of
untyped components.  Second, Natural's wrapper-based blame tracking is
more useful than Transient's ``collective blame'' tracking algorithm,
which in turn is superior to Erasure. {\em But\/}, the application of the
method also indicates problems with the expectations of theoreticians. The
first problem concerns the difference between the methods; neither is {\it
Natural\/} vastly more useful than {\it Transient\/} nor are the two
blame assignment methods highly superior to {\it Erasure\/}.  The second
one is that the cost of Transient's blame can be huge.  In turn, these
problems suggest that, on one hand, the existing theory does not predict
practice properly, and on the other hand, the existing practice may need
additional experiments to collect more data.

### 2021-03-05 writing/landscape.tex
issues with sections 9, 10, and 11, plus triggers for global searches
Fix citation style

The three blame strategies rely on three different ways of catching problems
with types at run-time: {\em Natural\/}, {\em Transient\/}, and {\em
Erasure\/}. This self-contained section summarizes these options with one illustrative example
using Racket syntax. The informed reader may wish to merely scan it.

@@

Under {\it the Natural semantics\/}, functions imported into and exported from
typed modules are wrapped in proxies that enforce the static type discipline
with run-time checks and track responsibilities~\cite{tf-popl-2008,
tfffgksst-snapl-2017}. Thus, when {\tt pack} is imported into a typed module,
the run-time system checks that it is a function and wraps it in a protective
proxy, which in turn, enforces the type of the function arguments with run-time checks.
Analogously the run-time system wraps each exported function of a typed module
(such as {\tt crypto-pack}) in a proxy that checks its arguments.  These checks
protect functions exported from typed modules against applications in untyped
code that supply the wrong kind of arguments.

@@

Under {\it the Transient semantics\/}, typed code is compiled so that all entry
points to functions check their arguments at run time and all function calls
check their return values against the expected type~\cite{vss-popl-2017}.
Furthermore, Transient uses \emph{shallow} checks, meaning they inspect only the
top constructor of a value. Since retrieving a value from within a structure (or
list, array, hash table etc.) is performed via a function call, the contents of a
complex value is checked on a piecemeal basis.

>>>>

The three blame strategies rely on three different ways of catching problems
with types at run-time: {\em Natural\/}, {\em Transient\/}, and {\em
Erasure\/}. This self-contained section summarizes these options with one
illustrative example using (Typed) Racket syntax. The informed reader may wish to merely
scan it.

@@

Under {\it the Natural semantics\/}, functions imported into and exported from
typed modules are wrapped in proxies that enforce the static type discipline
with run-time checks and track responsibilities~\citep{tf-popl-2008,
tfffgksst-snapl-2017}. Thus, when {\tt pack} is imported into a typed module,
the run-time system checks that it is a function and wraps it in a protective
proxy, which in turn, enforces the type of the function arguments with run-time checks.
Analogously the run-time system wraps each exported function of a typed module
(such as {\tt crypto-pack}) in a proxy that checks its arguments.  These checks
protect functions exported from typed modules against applications in untyped
code that supply the wrong kind of arguments.

@@

Under {\it the Transient semantics\/}, typed code is compiled so that all entry
points to functions check their arguments at run time and all function calls
check their return values against the expected type~\citep{vss-popl-2017}.
Furthermore, Transient uses \emph{shallow} checks, meaning they inspect only the
top constructor of a value. Since retrieving a value from within a structure (or
list, array, hash table etc.) is performed via a function call, the contents of a
complex value is checked on a piecemeal basis.

### 2021-03-05 writing/mutate-benchmarks.tex
Fix citation style

What exists is \citet{gtnffvf-jfp-2019}'s collection of Typed Racket programs
for systematically measuring the implementation's performance. The benchmark
suite consists of fully typed, correct programs, written by a number of
different authors, who have maintained and evolved these programs over several
years. The programs range widely in size, complexity, purpose, origin and in
programming style. The latter is mostly a reference to their reliance on many
Typed Racket features: occurrence typing~\cite{tf-icfp-2010}, types for mutable
and immutable data structures~\cite{hpst-sfp-2010}, types for first-class
classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's numeric
tower~\cite{stathff-padl-12}.

>>>>

What exists is \citet{gtnffvf-jfp-2019}'s collection of Typed Racket programs
for systematically measuring the implementation's performance. The benchmark
suite consists of fully typed, correct programs, written by a number of
different authors, who have maintained and evolved these programs over several
years. The programs range widely in size, complexity, purpose, origin and in
programming style. The latter is mostly a reference to their reliance on many
Typed Racket features: occurrence typing~\citep{tf-icfp-2010}, types for mutable
and immutable data structures~\citep{hpst-sfp-2010}, types for first-class
classes and objects~\citep{tsdtf-oopsla-2012}, and types for Racket's numeric
tower~\citep{stathff-padl-12}.

### 2021-03-05 writing/mutate-interesting.tex
Fix citation style

\item The goal of a comparative evaluation is to give the rational programmer a
chance to debug the same scenario using different pieces of information.  In the
case of gradual typing semantics, a meta-theorem due to Greenman and
Felleisen~\cite{gf-icfp-2018} says that if a program raises an exception under
Erasure, it also errors under all other semantics.  Hence, a comparison of blame
information insists that an interesting mutant {\em raises a run-time exception
under Erasure\/}.

>>>>

\item The goal of a comparative evaluation is to give the rational programmer a
chance to debug the same scenario using different pieces of information.  In the
case of gradual typing semantics, a meta-theorem due to Greenman and
Felleisen~\citep{gf-icfp-2018} says that if a program raises an exception under
Erasure, it also errors under all other semantics.  Hence, a comparison of blame
information insists that an interesting mutant {\em raises a run-time exception
under Erasure\/}.

### 2021-03-05 writing/mutate-mutators.tex
Fix citation style

Figure~\ref{table:mutation-ops} describes 16 mutators that satisfy these
constraints. As the last column indicates, some specialize or
generalize~\citet{lksfd-popl-2020}'s mutators, which in turn are borrowed from
the vast literature on mutation testing~\cite{jia2011analysis}.  Only two are
directly inherited; many mutators are brand new. For the latter, the authors
relied on their decades-long experience of making type-level mistakes in Typed
Racket, some of which take non-trivial effort to debug.

Most of the mutators are self-explanatory.  The first six apply to all
gradually typed languages; the next six to those that include classes and
objects. The last four target distinguishing features of Typed
Racket's type system,
specifically its sophisticated type system. For example, one mutation produced by {\tt arithmetic}
replaces a {\tt +} with a {\tt -} in an attempt to change the type of the
arithmetic expression; {\tt +}'s result is a {\tt Positive-Integer} when all
arguments are positive integers, while {\tt -} yields {\tt
Integer}~\cite{stathff-padl-12}. Similarly, the other three aim to confound
the occurrence type system.

>>>>

Figure~\ref{table:mutation-ops} describes 16 mutators that satisfy these
constraints. As the last column indicates, some specialize or
generalize~\citet{lksfd-popl-2020}'s mutators, which in turn are borrowed from
the vast literature on mutation testing~\citep{jia2011analysis}.  Only two are
directly inherited; many mutators are brand new. For the latter, the authors
relied on their decades-long experience of making type-level mistakes in Typed
Racket, some of which take non-trivial effort to debug.

Most of the mutators are self-explanatory.  The first six apply to all
gradually typed languages; the next six to those that include classes and
objects. The last four target distinguishing features of Typed
Racket's type system,
specifically its sophisticated type system. For example, one mutation produced by {\tt arithmetic}
replaces a {\tt +} with a {\tt -} in an attempt to change the type of the
arithmetic expression; {\tt +}'s result is a {\tt Positive-Integer} when all
arguments are positive integers, while {\tt -} yields {\tt
Integer}~\citep{stathff-padl-12}. Similarly, the other three aim to confound
the occurrence type system.

### 2021-03-05 writing/related.tex
Fix citation style

The literature on gradual typing presents many semantics beyond the three used
here and two additional blame strategies.  Pyret (pyret.org) assigns fixed-size
data types the Natural semantics and functions a Transient semantics. The
Forgetful~\cite{cl-icfp-2017} and the Amnesic~\cite{gfd-oopsla-2019} semantics
are similar to Transient but use wrappers instead of in-lined checks.
Nom~\cite{mt-oopsla-2017} and other \emph{concrete\/}
semantics~\cite{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
rat-oopsla-2017} assume that every value comes with a type tag and use tag
checks to supervise the interactions between typed and untyped code.  Monotonic
references~\cite{svctg-esop-2015} and the semantics~\cite{tlt-popl-2019,
etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
Gradual Typing technique~\cite{gct-popl-2016} are variants of Natural.  Among
these representative semantics, only Amnesic and Nom present innovative blame
strategies.  The experiment excludes these strategies because they seem to be
merely theoretical constructions and/or impose severe restrictions on
programmers.

>>>>

The literature on gradual typing presents many semantics beyond the three used
here and two additional blame strategies.  Pyret (pyret.org) assigns fixed-size
data types the Natural semantics and functions a Transient semantics. The
Forgetful~\citep{cl-icfp-2017} and the Amnesic~\citep{gfd-oopsla-2019} semantics
are similar to Transient but use wrappers instead of in-lined checks.
Nom~\citep{mt-oopsla-2017} and other \emph{concrete\/}
semantics~\citep{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
rat-oopsla-2017} assume that every value comes with a type tag and use tag
checks to supervise the interactions between typed and untyped code.  Monotonic
references~\citep{svctg-esop-2015} and the semantics~\citep{tlt-popl-2019,
etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.  Among
these representative semantics, only Amnesic and Nom present innovative blame
strategies.  The experiment excludes these strategies because they seem to be
merely theoretical constructions and/or impose severe restrictions on
programmers.

### 2021-03-05 writing/why-rational.tex
Fix citation style

Now consider the concrete (and simplistic) example of figure~\ref{fig:rational}.
Initially the code base consists of the two \dyncolor\ modules on the left plus
the \typecolor\ module at the bottom; \dyncolor\ indicates untyped, while
\typecolor\ means typed. When a typed module imports an untyped module, it must
assign types to the imported identifiers for the type checker's sake. Here {\tt
main} specifies that {\tt na-client} consumes a {\tt Real} number and produces a
non-positive one.\footnote{Typed Racket's type system reflects Racket's subset
reasoning about numbers, not (only) machine-level
representations~\cite{stathff-padl-12}.} A program execution ends in this error:
\begin{verbatim}
    na-client: broke its own contract
      promised: (<=/c 0)
      produced: 40
      in: (-> any/c (<=/c 0))           
      contract from: (interface for na-client)
      blaming: (interface for na-client)
       (assuming the contract is correct)
\end{verbatim}
The referenced contract is the compilation of the type of {\tt na-client}. The
definitive hint is ``{\tt blaming: (interface for na-client)}'' with the caveat
``{\tt (assuming the contract is correct)}.''

>>>>

Now consider the concrete (and simplistic) example of figure~\ref{fig:rational}.
Initially the code base consists of the two \dyncolor\ modules on the left plus
the \typecolor\ module at the bottom; \dyncolor\ indicates untyped, while
\typecolor\ means typed. When a typed module imports an untyped module, it must
assign types to the imported identifiers for the type checker's sake. Here {\tt
main} specifies that {\tt na-client} consumes a {\tt Real} number and produces a
non-positive one.\footnote{Typed Racket's type system reflects Racket's subset
reasoning about numbers, not (only) machine-level
representations~\citep{stathff-padl-12}.} A program execution ends in this error:
\begin{verbatim}
    na-client: broke its own contract
      promised: (<=/c 0)
      produced: 40
      in: (-> any/c (<=/c 0))           
      contract from: (interface for na-client)
      blaming: (interface for na-client)
       (assuming the contract is correct)
\end{verbatim}
The referenced contract is the compilation of the type of {\tt na-client}. The
definitive hint is ``{\tt blaming: (interface for na-client)}'' with the caveat
``{\tt (assuming the contract is correct)}.''
